---
title: "HW10"
output: html_document
date: "2026-05-06"
---

```{r}
library(tidyverse)
```
```{r}
# read sample names and extract
fam <- read_table("~/bioe-591-genomics/students/ashlynhemmah/HW10/species.int.fam", 
  col_names = FALSE,
  show_col_types = FALSE
)
samples <- fam$X2   # individual IDs are usually column 2

# choose your K value (will be 2 for this demo)
K <- 2

# read Q matrix
q <- read_table("~/bioe-591-genomics/students/ashlynhemmah/HW10/species.int.2.Q",
  col_names = FALSE,
  show_col_types = FALSE
)

# name the ancestry columns
colnames(q) <- paste0("Cluster", 1:K)

# combine with sample names
q_df <- q %>%
  mutate(sample = samples) %>%
  relocate(sample)

# convert to long format for ggplot
q_long <- q_df %>%
  pivot_longer(
    cols = starts_with("Cluster"),
    names_to = "cluster",
    values_to = "ancestry"
  ) %>%
  mutate(sample = factor(sample, levels = samples))
```

```{r}
# plot
ggplot(q_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color="white") +
  theme_bw() +
  labs(x = "Individual", y = "Ancestry proportion") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```


```{r}
cv_df <- tibble(file = "~/bioe-591-genomics/students/ashlynhemmah/HW10/admixture_AH_HW10_2-3934486.out") %>%
  mutate(
    text = map_chr(file, read_file),
    cv_line = str_extract(text, "CV error \\(K=\\d+\\):\\s*[-0-9.eE]+"),
    K  = str_match(cv_line, "CV error \\(K=(\\d+)\\)")[,2] |> as.integer(),
    CV = str_match(cv_line, ":\\s*([-0-9.eE]+)")[,2] |> as.numeric()
  ) %>%
  select(file, K, CV) %>%
  arrange(K)
cv_df
```

```{r}
library(adegenet)
library(vcfR)
```
```{r}
vcf <- read.vcfR(file = "~/bioe-591-genomics/course-materials/data/structure_data/squirrels.vcf", verbose = TRUE)
```
```{r}
dna <- vcfR2DNAbin(vcf, unphased_as_NA = F, consensus = T, extract.haps = F)
species_genind <- DNAbin2genind(dna)
species_genind
```
```{r}
species_genind_scaled <- scaleGen(species_genind,NA.method="mean",scale=F)
species_pca <- prcomp(species_genind_scaled, center=F,scale=F)
```

```{r}
screeplot(species_pca)
```
```{r}
pc <- data.frame(species_pca$x[,1:3])
pc$sample <- rownames(pc)
```

```{r}
ggplot(data=pc,aes(x=PC1,y=PC2))+
  geom_text(aes(label=sample))
```
```{r}
grp <- find.clusters(species_genind, n.pca = 50, n.clust = 2)
grp
```

```{r}
pc$cluster <- grp$grp
ggplot(pc, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2) +
  geom_text(aes(label = sample), vjust = -0.5, size = 3) 
```

```{r}
dapc1 <- dapc(species_genind, pop = grp$grp, n.pca = 50, n.da = 2)
dapc1
```

```{r}
q <- as.data.frame(dapc1$posterior)
  q$sample <- rownames(q)
  q_long <- q |>
    pivot_longer(
      cols = -sample,
      names_to = "cluster",
      values_to = "ancestry"
    )
q_long
```
```{r}
ggplot(q_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color = "white") +
  theme_bw() +
  labs(x = "Individual", y = "Assignment probability") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

```{r}
# identify clusters
grp_auto <- find.clusters(
  species_genind,
  n.pca = 50,
  choose.n.clust = FALSE, 
  max.n.clust = 10,
  stat = "BIC"
)

# print BIC values
grp_auto$Kstat
```

```{r}
# plot!
plot(
  1:length(grp_auto$Kstat),
  grp_auto$Kstat,
  type = "b",
  xlab = "K",
  ylab = "BIC"
)
```
**HOMEWORK**
```{r}
# read sample names and extract
fam1 <- read_table("species.int.fam", 
  col_names = FALSE,
  show_col_types = FALSE
)
samples <- fam1$X2   # individual IDs are usually column 2

# choose your K value (will be 2 for this demo)
K <- 2

# read Q matrix
q1 <- read_table("species.int.2.Q",
  col_names = FALSE,
  show_col_types = FALSE
)

# name the ancestry columns
colnames(q1) <- paste0("Cluster", 1:K)

# combine with sample names
q1_df <- q1 %>%
  mutate(sample = samples) %>%
  relocate(sample)

# convert to long format for ggplot
q1_long <- q1_df %>%
  pivot_longer(
    cols = starts_with("Cluster"),
    names_to = "cluster",
    values_to = "ancestry"
  ) %>%
  mutate(sample = factor(sample, levels = samples))
```


```{r}
# plot
ggplot(q1_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color="white") +
  theme_bw() +
  labs(x = "Individual", y = "Ancestry proportion") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

```{r}
# read sample names and extract
fam2 <- read_table("species.int.fam", 
  col_names = FALSE,
  show_col_types = FALSE
)
samples <- fam2$X2   # individual IDs are usually column 2

# choose your K value (will be 2 for this demo)
K <- 4

# read Q matrix
q2 <- read_table("species.int.4.Q",
  col_names = FALSE,
  show_col_types = FALSE
)

# name the ancestry columns
colnames(q2) <- paste0("Cluster", 1:K)

# combine with sample names
q2_df <- q2 %>%
  mutate(sample = samples) %>%
  relocate(sample)

# convert to long format for ggplot
q2_long <- q2_df %>%
  pivot_longer(
    cols = starts_with("Cluster"),
    names_to = "cluster",
    values_to = "ancestry"
  ) %>%
  mutate(sample = factor(sample, levels = samples))
```

```{r}
# plot
ggplot(q2_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color="white") +
  theme_bw() +
  labs(x = "Individual", y = "Ancestry proportion") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

```{r}
# read sample names and extract
fam3 <- read_table("species.int.fam", 
  col_names = FALSE,
  show_col_types = FALSE
)
samples <- fam3$X2   # individual IDs are usually column 2

# choose your K value (will be 2 for this demo)
K <- 6

# read Q matrix
q3 <- read_table("species.int.6.Q",
  col_names = FALSE,
  show_col_types = FALSE
)

# name the ancestry columns
colnames(q3) <- paste0("Cluster", 1:K)

# combine with sample names
q3_df <- q3 %>%
  mutate(sample = samples) %>%
  relocate(sample)

# convert to long format for ggplot
q3_long <- q3_df %>%
  pivot_longer(
    cols = starts_with("Cluster"),
    names_to = "cluster",
    values_to = "ancestry"
  ) %>%
  mutate(sample = factor(sample, levels = samples))
```

```{r}
# plot
ggplot(q3_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color="white") +
  theme_bw() +
  labs(x = "Individual", y = "Ancestry proportion") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

**The paper that this data came from concluded that K=2 is the best fit for the neutral dataset but for the northern population k=6 is highly biologically meaningful. In this analysis we can see that the K=2 is the best fit for the data in the PCA analysis there is a distance split between the two clusters but cluster two is a bit more spread out. Our graph of K=6 is less scattered than the K=4 graph but it still has about 3 individuals that don't fit into a single population. the elbow of our BIC plot is at a K=2 giving faurther evidence that it is the best fit for this data.**





